---
title: "Project 1"
author: "Name:  \n Partner: "
date: "`r Sys.Date()`"
output:
  html_document:
    toc: yes
    toc_depth: 3
  pdf_document:
    toc: yes
    toc_depth: '3'
header-includes:
- \usepackage{booktabs}
- \usepackage{longtable}
- \usepackage{array}
- \usepackage{multirow}
- \usepackage{wrapfig}
- \usepackage{float}
- \usepackage{colortbl}
- \usepackage{pdflscape}
- \usepackage{tabu}
- \usepackage{threeparttable}
- \usepackage{threeparttablex}
- \usepackage[normalem]{ulem}
- \usepackage{makecell}
- \usepackage{xcolor}
---
```{r setup, include = FALSE}
#### Load necessary packages ####
# * These packages are not necessary to complete the assignment and or only used 
#   to provide an example. 
packages <- c("knitr", "kableExtra", "magrittr", "readr", "geosphere", "dplyr")
install_me <- packages[!(packages %in% installed.packages()[, "Package"])]
if (length(install_me)) install.packages(install_me)
library(knitr)
library(kableExtra)
library(magrittr)
library(readr)
library(geosphere)
library(dplyr)
```

## Background
The World Health Organization has recently employed a new data science initiative, *CSIT-165*, that uses data science to characterize pandemic diseases. 
*CSIT-165* disseminates data driven analyses to global decision makers.

*CSIT-165* is a conglomerate comprised of two fabricated entities: *Global Health Union (GHU)* and *Private Diagnostic Laboratories (PDL)*. 
Your and your partner's role is to play a data scientist from one of these two entities.

## Data
> [2019 Novel Coronavirus COVID-19 (2019-nCoV) Data Repository by John Hopkins CSSE](https://github.com/CSSEGISandData/COVID-19/tree/master/csse_covid_19_data/csse_covid_19_time_series)
Data for 2019 Novel Coronavirus is operated by the John Hopkins University Center for Systems Science and Engineering (JHU CSSE).
Data includes daily time series CSV summary tables, including confirmations, recoveries, and deaths. 
Country/region are countries/regions hat conform to World Health Organization (WHO).
Lat and Long refer to coordinates references for the user. 
Date fields are stored in MM/DD/YYYY format.

## Project Objectives

### Objective 1: What was the origin country of the COVID-19 outbreak?
```{r ob1}
#load data
confirmed_cases <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_confirmed_global.csv"
covid_deaths <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_deaths_global.csv"

cases_df <- read.csv(confirmed_cases, header = TRUE, na.strings = c("", " "))
deaths_df <- read.csv(covid_deaths, header = TRUE, na.strings = c("", " "))

#segment first day of COVID data
data_cases <- dplyr::select(cases_df, Province.State, Country.Region, X1.22.20); 
data_deaths <- dplyr::select(deaths_df, Province.State, Country.Region, X1.22.20)

# Filter for the first day and select relevant columns
first_day_cases <- cases_df %>%
  filter(X1.22.20 != 0) %>%
  select(Province.State, Country.Region, X1.22.20)
first_day_deaths <- deaths_df %>%
  filter(X1.22.20 != 0) %>%
  select(Province.State, Country.Region, X1.22.20)

# Identify the area with the highest confirmed cases and highest deaths
max_cases <- first_day_cases %>%
  filter(X1.22.20 == max(X1.22.20)) %>%
  pull(Province.State)
max_deaths <- first_day_deaths %>%
  filter(X1.22.20 == max(X1.22.20)) %>%
  pull(Province.State)

# Determine if the area(s) identified is the origin of the outbreak
if(max_cases == max_deaths) {
  output <- paste("The origin of the COVID-19 outbreak was likely", max_cases)
  print(output)
}
```

### Objective 2: Where is the most recent area to have a first confirmed case?

```{r ob2}

# iterates through each (date-containing) column 
for(date_column in (5:ncol(cases_df))){
  
  # iterates through each row (case count) for that specific date
  for(x in (1:length(cases_df[,date_column]))){ # subsets the column for a single date
    if(cases_df[x, date_column] == 1 & cases_df[x, date_column-1] == 0){ # checks if there is a new case 
      newest_case <- cases_df[x, 2] # updates variable with the corresponding country name (column 2) 
    }
  }
}

cat("The most recent area to have a first confirmed case is", newest_case)

```

### Objective 3: How far away are the areas from objective 2 from where the first confirmed case(s) occurred? 
```{r ob3}
# assigns correct values to origin country and recent country variables using output from ob1 and ob2
recent_region <- newest_case
origin_city <- max_cases
origin_country <- "China"

# subsets the lat and long values, using the row number that corresponds to the origin country
origin_lat = cases_df[which(cases_df$Province.State == origin_city), 3]
origin_long = cases_df[which(cases_df$Province.State == origin_city), 4]
# creates a list of (longitude, latitude) -> the correct input format for distm
origin_coordinates <- c(origin_long, origin_lat)

# subsets the lat and long values, using the row number that corresponds to the most recent country
recent_lat = cases_df[which(cases_df$Country.Region == recent_region), 3]
recent_long = cases_df[which(cases_df$Country.Region == recent_region), 4]
# creates a list of (longitude, latitude) -> the correct input format for distm
recent_coordinates <- c(recent_long, recent_lat)

# calculates distance between coordinates using distm
distance = distm(origin_coordinates, recent_coordinates, fun=distGeo)
# converts dist in meters to miles using conversion factor
miles_distance = distance/1609

# prints distance between two locations in a sentence using values from above
sprintf("%s is %f miles away from %s, %s", recent_region, miles_distance, origin_city, origin_country)




```

### Objective 4

#### Objective 4.1
```{r ob4.1}
#filter out cruise ship data from the data sets
cases_df <- cases_df[!(is.na(cases_df$Lat) | cases_df$Lat == 0),]
deaths_df <- deaths_df[!(is.na(deaths_df$Lat) | deaths_df$Lat == 0),]

# Extract the most recent date
most_recent_date <- tail(colnames(cases_df), 1)
print(most_recent_date)


# Extract the columns for the most recent date
confirmed_cases_latest<- dplyr::select(cases_df, Province.State, Country.Region, ncol(cases_df)); 
deaths_latest  <- dplyr::select(deaths_df, Province.State, Country.Region, ncol(deaths_df))

#merge the data sets into one
merged_df <- merge(confirmed_cases_latest, deaths_latest, by = c("Province.State", "Country.Region"))

# Calculate risk scores for the most recent date
confirmed_cases_latest <- "X3.9.23.x"
deaths_latest <- "X3.9.23.y"
risk_scores_latest <- merged_df[,deaths_latest] / merged_df[,confirmed_cases_latest] * 100
#risk_scores_latest <- #as.numeric(as.matrix(risk_scores_latest)))
# Add risk scores to the end of the table as a new column
mutate(merged_df, risk_score = risk_scores_latest)
head(merged_df)

```
```
# Find the minimum risk score across all provinces for the most recent date
min_risk_score_latest <- min(risk_scores_latest, na.rm = TRUE)
print(min_risk_score_latest)

<<<<<<< HEAD
#Find the provinces with the minimum risk score for the most recent date
min_risk_provinces <- risk_scores_latest
head( min_risk_provinces)
```
# Where I'm stuck
=======
# Find the provinces with the minimum risk score for the most recent date
min_risk_provinces_latest <- colnames(risk_scores_latest)[which(risk_scores_latest == min_risk_score_latest, arr.ind = TRUE)[, "col"]]
head( min_risk_provinces_latest)
>>>>>>> acc5ccea2a25942a82b7dcc157b01e2d852879eb

# Filter cases_df to only include the provinces with the minimum risk score for the most recent date
min_risk_cases_latest <- cases_df[, c(1:4, min_risk_provinces_latest, most_recent_date)]

# Find the province with the most confirmed cases among those with the minimum risk score for the most recent date
max_cases_province_latest <- min_risk_cases_latest[which.max(min_risk_cases_latest[, ncol(min_risk_cases_latest) - 1]), "Province.State"]

# Extract the confirmed cases number for the province with the lowest risk score and the most confirmed cases for the most recent date
max_cases_confirmed_latest <- min_risk_cases_latest[which(min_risk_cases_latest[, "Province.State"] == max_cases_province_latest), most_recent_date]

```
# Filter cases_df to only include the provinces with the minimum risk score for the most recent date
min_risk_cases_latest <- cases_df[, c(1:4, min_risk_provinces_latest, most_recent_date)]
```
# Find the province with the most confirmed cases among those with the minimum risk score for the most recent date
max_cases_province_latest <- min_risk_cases_latest[which.max(min_risk_cases_latest[, ncol(min_risk_cases_latest) - 1]), "Province.State"]

# Extract the confirmed cases number for the province with the lowest risk score and the most confirmed cases for the most recent date
max_cases_confirmed_latest <- min_risk_cases_latest[which(min_risk_cases_latest[, "Province.State"] == max_cases_province_latest), most_recent_date]

# Output the result
paste("The province with the lowest risk score and the most confirmed cases on", most_recent_date, "is", max_cases_province_latest, "with a risk score of", min_risk_score_latest, "and", max_cases_confirmed_latest, "confirmed cases.")





# Store location columns in separate variables
location_cols_cases <- cases_df[, 1:4]
location_cols_deaths <- deaths_df[, 1:4]

# Remove location columns from data frames
cases_df <- cases_df[, -c(1:4)]
deaths_df <- deaths_df[, -c(1:4)]

# Calculate risk scores
risk_scores <- deaths_df / cases_df * 100

# Add location columns back to risk_scores data frame
risk_scores <- cbind(location_cols_deaths, risk_scores)

# Calculate risk scores
risk_scores <- deaths_df[, -c(1:4)] / cases_df[, -c(1:4)] * 100

#Find the minimum risk score across all provinces
min_risk_score <- min(risk_scores, na.rm = TRUE)
print(min_risk_score)

min_risk <- which(risk_scores == min_risk_score)
min_risk_provinces <- risk_scores$Province.State[min_risk]
print(min_risk_provinces)

# Find the provinces with the minimum risk score
min_risk_provinces <- colnames(risk_scores)[which(risk_scores == min_risk_score, arr.ind = TRUE)[, "col"]]

# Filter cases_df to only include the provinces with the minimum risk score
min_risk_cases <- cases_df[, c(1:4, min_risk_provinces)]

# Find the province with the most confirmed cases among those with the minimum risk score
max_cases_province <- min_risk_cases[which.max(min_risk_cases[, ncol(min_risk_cases)]), "Province.State"]

# Output the result
paste("The province with the lowest risk score and the most confirmed cases is", max_cases_province, "with a risk score of", min_risk_score)

```

#### Objective 4.2
```{r ob4.2}
```

### Objective 5
```{r ob5}
# creates a list of all countries by subsetting the country column of the dataset
countries <- cases_df$Country.Region
# removes duplicates (due to multiple provinces)
countries <- unique(countries)

deaths = 0
cases = 0

# initialization of empty lists
country_cases <- c()
country_deaths <- c()

# iterates through each unique country
for(country in countries){
  # creates a list of the indexes of every time the country appears 
  country_duplicates <- which(cases_df$Country.Region == country)
  # iterates through each of the duplicate indexes (each province of that country)
  for(dup in country_duplicates){
    # adds the cases for that province to the total count for the country
    cases <- cases + cases_df[dup, 1147]
  }
  # appends the total case count for that country to the country_cases list
  country_cases <- append(country_cases, cases)
  # resets case count to 0 before moving on to the next country in the list
  cases = 0
}

# repeats same procedure, but with deaths data set
for(country in countries){
  country_duplicates <- which(deaths_df$Country.Region == country)
  for(dup in country_duplicates){
      deaths <- deaths + deaths_df[dup, 1147]
  }
  country_deaths <- append(country_deaths, deaths)
  deaths = 0
}

# creates a data frame with countries, their respective cases, and respective deaths
overview <- data.frame(countries, country_cases, country_deaths)

# uses arrange function to create two different data frames, each sorted in descending order of case/death
casewise <- arrange(overview, -country_cases)
deathwise <- arrange(overview, -country_deaths)

# creates new data frames that subset only the top 5 
top_case <- casewise[1:6,]
top_death <- deathwise[1:6,]

# uses kable to display the data frames as visual tables
kable(top_case)
kable(top_death)


```

### GitHub Log
```{bash gitlog} 
git log --pretty=format:"%nSubject: %s%nAuthor: %aN%nDate: %aD%nBody: %b"
```
